---
title: "Clustering Spectra from High Resolution DI-MS/MS Data Using CluMSID"
author: "Tobias Depke"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{CluMSID DI-MS/MS Tutorial}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignetteDepends{CluMSIDdata}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    echo = TRUE,
    warning = FALSE
)
```

<style>
p.caption {
    font-size: 80%;
    text-align: left;
}
</style>

```{r captions, include=FALSE}
fig1 <- paste("**Figure 1:**",
                "Circularised dendrogram as a result of",
                "agglomerative hierarchical clustering with average linkage",
                "as agglomeration criterion based on",
                "MS^2^ spectra similarities",
                "of the DI-MS/MS example data set.",
                "Each leaf represents one feature and colours encode",
                "cluster affiliation of the features.",
                "Leaf labels display feature IDs, along with",
                "feature annotations, if existent.",
                "Distance from the central point is indicative",
                "of the height of the dendrogram.")
```

## Introduction

Although originally developed for liquid chromatography-tandem mass 
spectrometry (LC-MS/MS) data, CluMSID can also be used with 
direct infusion-tandem mass spectrometry (DI-MS/MS) data.

Generally, the missing retention time dimension makes feature annotation
in metabolomics harder but if only direct infusion data is at hand,
CluMSID can help to get an overview of the chemodiversity of a sample
measured by DI-MS/MS.

In this example, we will use a similar sample (1uL *Pseudomonas aeruginosa* 
PA14 cell extract) as in the General Tutorial, measured on the same machine,
a Bruker maxis^HD^ qTOF operated in ESI-(+) mode with auto-MS/MS but without
chromatographic separation.

## Data import

We load the file from the `CluMSIDdata` package:

```{r packages, message=FALSE, warning=FALSE}
library(CluMSID)
library(CluMSIDdata)

DIfile <- system.file("extdata", 
                        "PA14_maxis_DI.mzXML",
                        package = "CluMSIDdata")
```

## Data preprocessing

The extraction of spectra works the same way as with LC-MS/MS data:

```{r extract}
ms2list <- extractMS2spectra(DIfile)
length(ms2list)
```

Merging of redundant spectra is less straightforward when retention time 
is not available. Depending on the MS/MS method it can be next to impossible 
to decide whether two spectra with the same precursor *m/z* and similar
fragmentation patterns derive from the same analyte or from two different 
but structurally similar ones.

In this example, we would like to merge spectra with identical precursor ions
only if they were recorded one right after another. We can do so by setting
`rt_tolerance` to 1 second:

```{r merge}
featlist <- mergeMS2spectra(ms2list, rt_tolerance = 1)
length(featlist)
```

We see that we have hardly reduced the number of spectra in the list.
If we would decide to merge all spectra with identical precursor *m/z*
from the entire run, we could do so by setting `rt_tolerance` to the
duration of the run, in this case approx. 250 seconds:

```{r merge2}
testlist <- mergeMS2spectra(ms2list, rt_tolerance = 250)
length(testlist)
```

The resulting number of spectra is drastically lower but the danger of
merging spectra that do not actually derive from the same analyte
is also very big.

## Generation of distance matrix

In this very explorative example, we skip the integration of previous
knowledge on feature identities and generate a distance matrix
right away:

```{r distmat, eval=FALSE}
distmat <- distanceMatrix(featlist)
```

```{r distmat2, include=FALSE}
load(file = system.file("extdata", "di-distmat.RData", package = "CluMSIDdata"))
```

## Data exploration

Starting from this distance matrix, we can use all the data exploration 
functions that `CluMSID` offers. In this example workflow, 
we look at a cluster dendrogram:

```{r dendrogram, fig.width=12, fig.asp=1, fig.cap=fig1}
CluMSID_HCplot(distmat)
```

It is directly obvious that we have some spectra that are nearly identical
and thus most likely derive from the same analyte, e.g. the many spectra with
a precursor *m/z* of 270.19. But we still see nice clustering of similar
spectra with different precursor *m/z*, e.g. the huge gray cluster that 
contains a lot of different alkylquinolone type metabolites (see General
Tutorial).

In conclusion, CluMSID is very useful to provide an overview of spectral
similarities within DI-MS/MS runs but wherever annotation is in the focus,
one should not do without the additional layer of information created by
chromatographic separation.

# Session Info
```{r session}
sessionInfo()
```
